\documentclass[12pt]{article}
\usepackage{sydewkrpt}
\usepackage{longtable}
\usepackage{array}
\usepackage{ragged2e}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{siunitx}
\DeclareMathOperator*{\argmin}{\arg\!\min}
\DeclareMathOperator*{\argmax}{\arg\!\max}
\newcolumntype{P}[1]{>{\RaggedRight\hspace{0pt}}p{#1}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%    Begin Document    %%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}
\pagenumbering{roman}

\waterlootitle{SYDE 531: Final Project Report}{
  Betting Strategies for an NCAA March Madness Tournament Bracket Using Dynamic Modern Portfolio Optimization\\
}{
  Brian Sinclair -- 20346309\\
  Riley Donelson -- 20342815\\
  }

\dotableofcontents

\newpage
\doublespacing
\pagenumbering{arabic}
\section{Introduction}
\setlength{\parindent}{1cm}
\subsection{Background}
This report details the design optimization of betting strategies for the NCAA Men's Division 1 Basketball Championship, commonly referred to as the ``March Madness'' tournament.
This tournament is a yearly event comprising of the top 64 teams in US College Basketball \cite{doi:10.1080/00031305.1996.10473540}.
It is structured in a classical tournament bracket format, beginning with 32 games (64 teams) and ending with one final Championship game.
The losing team from each game is eliminated from the tournament at each round, which cuts the number of games at each successive round in half.
This provides a tournament layout consisting of 6 rounds.

The attempt at predicting and/or betting on the games in this tournament is a common activity for many fans and followers of the sport.
In order to provide more informed decisions, the design optimization of betting strategies under probabilistic uncertainty was conducted using a modern portfolio theory mean-variance approach, combined with dynamic programming to select the optimal traversal of bet placements through the tournament.
The formulation of each optimization algorithm is provided, as well as the results and analysis of the solution.
Discussions on the results, sensitivity of the system, and conclusions are also provided. 

\subsection{Problem Definition}


\newpage
\section{Optimization Algorithms}
\subsection{Modern Portfolio Optimization}
\subsubsection{Approach Background}
As previously mentioned the goal of the project is to maximize the return on investment when betting on individual games in the yearly NCAA March Madness Tournament.
It was recognized early on that gambling in sports follows a similar behaviour to that of portfolio optimization between two assets. 
Any one game can be thought of as a portfolio where each asset is team A, or team B in the match-up.
The objective of any one game (as a portfolio) is to maximize return on investment when given a budget to spend on that game.
For the purposes of the tournament, instead of the portfolio being an individual game, it becomes an individual round (ie. Portfolio's 1 through 6 are Rounds 1 through 6).
The portfolio then changes from being constructed of assets as teams, to assets as each game.
In other words, one game is one asset in the portfolio (Round).
This outlook would follow a myopic outlook on Portfolio Theory for each round.
In the first round there is then 32 assets to purchase, 16 in the second, 8 in the third, and so on.

The bettor, or owner of this newly defined portfolio then has the option to buy into a game (wager a percentage of their budget) on either team A, or team B in an (A,B) match-up. 
If the wager is a positive value, then the bettor is placing a wager on team A.
If it is negative, then the bettor is placing a wager on team B.
The myopic optimization problem is to maximize the return on investment in any one round by maximizing budget allocation across all assets. 
The following section will aim to describe a novel application of expanding on the myopic approach into a dynamic portfolio optimization problem where the budget for the following round is the return on investment from the previous round.


\subsubsection{Setup}
Much of the general setup and background was already previously described. 
As previously mentioned the objective is to use modern portfolio theory techniques to determine a maximum return on investment for an individual round basis.
This is done using a mean-variance optimization approach described below.

\begin{equation*}
\begin{aligned}
& \underset{x}{\text{minimize}}
& & f(x) =  E(R^Tx)-\theta\sqrt{var(R^Tx)}\\
& \text{s.t.}
& & \sum{x}=b\\
&&& |x_i| \leq b, \; i = 1, \ldots, m.
\end{aligned}
\end{equation*}
where,

\begin{table}[htbp!]
\begin{centering}
    \begin{tabular}{r l}
    $x_i:$        & Quantity wagered on a single game \\ 
    $R:$        & Vector of expected returns             \\ 
    $b:$        & Budget for any one round            \\ 
    $m:$        & Total Number of games(assets)    \\ 
    \end{tabular}
    \label{dp_vars}
    \caption{Variables used in the Dynamic Programming formulation.}
\end{centering}
\end{table}

Probabilities -- and how they are calculated

\subsubsection{Assumptions}


Uniform Distribution

\subsection{Dynamic Programming}
The method of discrete dynamic programming was employed in this solution to take the values provided from the mean-variance approach to portfolio optimization, and select the best ``path'' through the tournament where bets should be placed to maximize return.
The betting possibilities are discretized into 5 states, wherein one may utilize 100\%, 75\%, 50\%, 25\%, or 0\% of their budget on a round of the tournament.

Using the discrete dynamic programming methodology outlined in class, the problem was divided into stages, with each stage representing a round of the tournament.
This method depends on Bellman's principle of optimality, stated as follows:
\begin{quote}
\emph{``The optimal set of decisions in a multistage process has the property that whatever the initial stage, state and decisions are, the remaining decisions, given the current stage, must form a set of optimal decisions for the remaining problem.''}
\end{quote}
This gives rise to the recursive nature of the problem formulation.
A breakdown of the variables used in the analysis, and their meaning, is provided in Table \ref{dp_vars}. \\

\begin{table}[htbp!]
\begin{centering}
    \begin{tabular}{|l|l|}
    \hline
    Variable & Meaning           \\ \hline
    $x_{t}$        & Decision Variable \\ \hline
    $s_{t}$        & State Variable    \\ \hline
    $t$        & Stage Number             \\ \hline
    $f$        & Return            \\ \hline
    $f^{*}$       & Maximum Return    \\ \hline
    \end{tabular}
    \label{dp_vars}
    \caption{Variables used in the Dynamic Programming formulation.}
\end{centering}
\end{table}

A decision (the value of $x_{t}$) is required at each stage, where $t$ is the stage, and 1 $\le t \le$ 6.
Each stage has a state variable, $s_{t}$, and this state is divided into five states, one for each allowed percentage of the total budget to be bet.
At each stage (i.e. each round in the tournament), the dynamic programming system moves from the current state in the current stage to another ideal feasible state in the next stage.
This discrete dynamic programming methodology is able to solve each stage problem individually by considering the return function for the current stage, and the optimal function, $f^{*}$, from the next stage ($t + 1$) to the end of the horizon (6 rounds).
In order to accomplish this, a recursive equation was constructed, as shown in Equation \ref{dp_recursive}.

\begin{equation}
f_{t}(s_{t}(i)) = \max \ [r_{t}(x_{t},s_{t}(i)) + f^{*}_{t+1}(s_{t+1}(j))]
\label{dp_recursive}
\end{equation}

This way, the system is able to determine optimal decisions for the problem overall, from any given state.
The full list of decisions and their individual returns are available at the end of the procedure.

Upon formulation of the problem, the dynamic programming portion of the system was then implemented in Matlab to generate solutions, given inputs from the prior mean-variance method.
The full implementation is given in Appendix A.
The data from the mean-variance solution was structured into a 5x5x6 3-dimensional matrix, where each row represents the returns of each state transition to the next stage.
Each ``page'' of this matrix (i.e. along the 3rd dimension) represents the data for each stage, or round, of the tournament.
With the data modelled appropriately, the optimal return (and its location) for the final round of the tournament was found and placed into a list holding each of the optimal decisions throughout the tournament.
Then, each stage of the tournament was looped through in a backwards fashion, and the optimal path of bets through the tournament was generated recursively.
The returns from each stage are then summed, to provide a look at the total optimal return for the tournament.

\subsubsection{Assumptions}
One assumption that needs to be made in calculating and implementing a dynamic programming graph traversal is that probabilities of future rounds do not change.
The algorithm finds an optimal path of discretized wagers progressing through the six rounds of the tournament. 
All benefits from from one round to another are calculated based off of expected return in the following round.
In order to calculate expected return, match-ups for all rounds of the tournament would need to be known, and correspondingly, the odds of each team is all match-ups. 
For the purposes of this project, match-ups from a past tournament was used to have knowledge of these match-ups, however this would not be realistic.
However this does allow for accurate knowledge of probabilities of every match-up in all rounds of the tournament and is necessary for implementing dynamic programming.

This could be overcome by using an intelligent system to predict probabilities of future match-ups and dynamically change algorithm as tournament progresses. 
For example, if round 1 of the tournament is complete and teams in round 2 were not predicted correctly then would need to dynamically change betting towards teams in current round (round 2).
The machine learning algorithm would then adopt a new prediction for rounds 3 through 6 and the entire betting path would change.
One solution to an intelligent system would be to implement an artificial neural network to use backpropagation techniques from past data to predict the outcome of a tournament. 
The team is working on this out of interest but it is not discussed further in this report.

\section{Results and Analysis}
\subsection{Sample Inputs, Outputs, and Detailed Solution}


\subsection{Sensitivity Analysis}


\subsection{Inferences on Problem Solutions}

\newpage
\section{Recommendations \& Conclusions}
Two Stage Stochastic Optimization

\newpage
\section{References}
\bibliographystyle{IEEEtran}
\bibliography{bib}

\newpage
\section{Appendix A: Full Implementation}
\subsection{Dynamic Programming}
\begin{verbatim}
clear;

num_stages = 6;	% number of rounds in the tournament
num_states = 5;	% number of possible bets at each round

opt_list = [];	% stores optimal return and location of bet
[state_benefit, opt_dsg] = max(benefits(:, :, num_stages));
[row, col] = max(state_benefit);
state_benefit = row;
opt_dsg = opt_dsg(col);

opt_list = [opt_list; state_benefit, opt_dsg];

for curr_stage = (num_stages-1):-1:1
	[state_benefit, opt_dsg] = max(benefits(:, opt_dsg, curr_stage));
	opt_list = [opt_list; state_benefit, opt_dsg];
end

opt_list
\end{verbatim}


\end{document}